### 2015-05-07 papers/histogram/thesis/thesis.tex
added a sample figure and an overview of the methods section

\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{figs/dos-sample.pdf}
  \caption[Sample density of states]{Sample density of states for
    square-well fluid with a well width $\lambda=1.3$, filling
    fraction $\eta=0.3$, and $N=20$ spheres, as computed by several
    histogram methods. The density of states has been normalized to
    have a maximum value of $1$.}
  \label{fig:dos_sample}
\end{figure}

@@

Having covered the basics of the square-well fluid as well as
Metropolis and biased Monte Carlo simulations, we can now discuss
several histogram methods for constructing energy weights.\fixme{more
  overview of the methods section?}

@@

\begin{algorithm}[!b]
  \caption{The simple flat method}
  \label{alg:simple_flat}
  \begin{alg}

@@

\fixme{find more sources to cite for some of the claims made in this
  section}

@@

The proportionality $\tilde T_{ij}\propto T_{ij}$ can be made explicit
by enforcing that the probabilities of all transitions from an energy
$E_j$ sum to unity, i.e.
\begin{align}
  \sum_iP\p{E_j\to E_i}=\sum_iT_{ij}=1,
  \label{eq:transition_norm_condition}
\end{align}
which implies
\begin{align}
  T_{ij}=\f{\tilde T_{ij}}{\sum_i\tilde T_{ij}}.
  \label{eq:transition_normalization}
\end{align}
To compute $T$ from $\tilde T$, one must therefore normalize each row
of $\tilde T$ independently.

@@

\begin{algorithm}[!b]
  \caption{Transition matrix Monte Carlo initialization}
  \label{alg:tmmc}
  \begin{alg}

@@

Figure \ref{fig:transitions_sample} provides an example of a
transition matrix $T_d\p{E,\Delta E}$ for a particular square-well
fluid. The yellow band at $\Delta E=0$ indicates that most moves
transition from an energy to itself. Low energies are more likely to
transition up in energy, with $\Delta E>0$, than down in energy, with
$\Delta E<0$. The high energies which are more likely to transition
down in energy than up are those above the state of maximum entropy.

\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{figs/transitions-sample.pdf}
  \caption[Sample transition matrix]{Sample transition matrix
    $T_d\p{E,\Delta E}$ for square-well fluid with a well width
    $\lambda=1.3$, filling fraction $\eta=0.3$, and $N=20$ spheres.}
  \label{fig:transitions_sample}
\end{figure}

@@

Once we identify an energy $E_-$ satisfying (\ref{eq:min_E_condition})
in an initialization routine for any histogram method, we effectively
forbid transitions to energies $E<E_-$ by tentatively setting weights
$w\p{E<E_-}=e^{-\p{E-E_-}/kT}$ for some temperature $T\ll
T_{\t{min}}$. These weights prevent the initialization routine from
wasting time on energies at which we will override the weight array
after initialization anyways. A special case was made for Wang-Landau,
as the design of this method assumes knowledge of the energy range of
interest a priori. In order to fairly compare Wang-Landau to other
methods, we therefore identified $E_-$ for each simulated system using
other methods, and hard-coded the minimum important energies for each
system into the Wang-Landau initialization routines\fixme{talk about
  how bad this is}. We always set
$w\p{E<E_-}=e^{-\p{E-E_-}/kT_{\t{min}}}$ after initialization of all
histogram methods with $kT_{\t{min}}=0.2\epsilon$. We chose this
temperature because it seems to be right below the temperature at
which the square-well fluid condenses from a gas into a
liquid (see Section \ref{sec:errors}). \fixme{what about pressure?}

@@

@@

\begin{figure}[H]
  \centering
  \caption[Energy histograms]{Energy histograms}
  \label{fig:histograms}
\end{figure}

\begin{figure}[H]
  \centering
  \caption[Density of states]{Density of states}
  \label{fig:density_of_states}
\end{figure}

\begin{figure}[H]
  \centering
  \caption[Energy weights]{Energy weights}
  \label{fig:weights}
\end{figure}

>>>>

\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{figs/dos-sample.pdf}
  \caption[Density of states]{Sample density of states for square-well
    fluid with a well width $\lambda=1.3$, filling fraction
    $\eta=0.3$, and $N=20$ spheres, as computed by several histogram
    methods. The density of states has been normalized to have a
    maximum value of $1$.}
  \label{fig:dos_sample}
\end{figure}

@@

Having covered the basics of the square-well fluid as well as
Metropolis and biased Monte Carlo simulations, we can now discuss
several histogram methods for constructing energy weights. In Section
\ref{sec:flat_histogram}, we will introduce the flat histogram method,
which is not directly applicable to the study of most systems, but
which forms the basis of many other methods. We will then introduce,
in Section \ref{sec:simple_flat}, the ``simple flat'' method, which we
designed to be the simplest viable histogram method, and against which
we can compare more complicated methods. Next, in Sections
\ref{sec:wang_landau}--\ref{sec:optimized_ensemble} we introduce three
published methods: Wang-Landau, transition matrix Monte Carlo, and the
optimized ensemble. The last of these methods we will not actually
implement and analyze because, unlike most histogram methods, it was
designed to optimize weights rather than construct them. We will,
however, borrow ideas from the optimized ensemble to design a
histogram method of our own, a hybrid of the optimized ensemble and
transition matrix Monte Carlo, in Section \ref{sec:oetmmc}. Finally,
we will cover some general implementation details and loose ends in
Section \ref{sec:loose_ends}

@@

\begin{algorithm}[tb]
  \caption{The simple flat method}
  \label{alg:simple_flat}
  \begin{alg}

@@

@@

The proportionality $\tilde T_{ij}\propto T_{ij}$ can be made explicit
by enforcing that the probabilities of all transitions from an energy
$E_j$ sum to unity, i.e.
\begin{align}
  \sum_iP\p{E_j\to E_i}=\sum_iT_{ij}=1,
  \label{eq:transition_norm_condition}
\end{align}
which implies
\begin{align}
  T_{ij}=\f{\tilde T_{ij}}{\sum_i\tilde T_{ij}}.
  \label{eq:transition_normalization}
\end{align}
Computing $T$ from $\tilde T$ therefore requires normalizing each row
of $\tilde T$ independently.

@@

\begin{algorithm}[tb]
  \caption{Transition matrix Monte Carlo initialization}
  \label{alg:tmmc}
  \begin{alg}

@@

Figure \ref{fig:transitions_sample} provides an example of a
transition matrix $T_d\p{E,\Delta E}$ for a particular square-well
fluid. The band at $\Delta E=0$ indicates that most moves transition
from an energy to itself. Low energies are more likely to transition
up in energy, with $\Delta E>0$, than down in energy, with $\Delta
E<0$. The high energies which are more likely to transition down in
energy than up are those above the state of maximum entropy.

\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{figs/transitions-sample.pdf}
  \caption[Transition matrix]{Sample transition matrix $T_d\p{E,\Delta
      E}$ for square-well fluid with a well width $\lambda=1.3$,
    filling fraction $\eta=0.3$, and $N=20$ spheres.  The band at
    $\Delta E=0$ indicates that any given state is most likely to
    transition into another state with the same energy.}
  \label{fig:transitions_sample}
\end{figure}

@@

Once we identify an energy $E_-$ satisfying (\ref{eq:min_E_condition})
in an initialization routine for any histogram method, we effectively
forbid transitions to energies $E<E_-$ by tentatively setting weights
$w\p{E<E_-}=e^{-\p{E-E_-}/kT}$ for some temperature $T\ll
T_{\t{min}}$. These weights prevent the initialization routine from
wasting time on energies at which we will override the weight array
after initialization anyways. A special case was made for Wang-Landau,
as the design of this method assumes knowledge of the energy range of
interest a priori. In order to fairly compare Wang-Landau to other
methods, we therefore identified $E_-$ for each simulated system using
other methods, and hard-coded the minimum important energies for each
system into the Wang-Landau initialization routines\fixme{talk about
  how bad this is}. We always set
$w\p{E<E_-}=e^{-\p{E-E_-}/kT_{\t{min}}}$ after initialization of all
histogram methods with $kT_{\t{min}}=0.2\epsilon$. We chose this
temperature because it is right below the temperature at which the
square-well fluid condenses from a gas into a liquid (see Section
\ref{sec:errors}). \fixme{what about pressure?}

@@

Having developed and implemented all of the histogram methods
discussed in Section \ref{sec:methods}, we will now look at the
results from simulating with these methods.

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{figs/array-samples.pdf}
  \caption[Sample arrays]{Sample arrays computed by various histogram
    methods for a square-well fluid with a well width $\lambda=1.3$,
    filling fraction $\eta=0.3$, and $N=20$ spheres after simulating
    for $10^5N^2$ iterations: (top left) energy histogram, (top right)
    weight array, (bottom left) density of states.}
  \label{fig:arrays}
\end{figure}
